---
title: "Virtual Portfolio 1"
author: "Tanvi Suresh"
output: 
  html_document:
    number_sections: yes
    toc: yes
    toc_float: yes 
          
date: "2022-09-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(tidyverse,sf,tmap,rio,dplyr)
```

## Importing and joining datasets
### Importing tracts geopackage and life expectancy csv file. Transforming GA Tracts to Albers Equal Area
```{r, message=F}

gatracts <- st_read('ga-tracts.gpkg') %>%
  st_transform(5070)
gale <- read.csv('GA_A (1).csv')

```

### Creating "tracts" dataset by merging GA life expectancy ("gale") dataset onto "gatracts" dataset with a leftjoin to keep the total number of tracts (even the tracts without life expectancy data -192 missing) 
```{r}
gale <- gale %>%
    mutate(Tract.ID=as.character(Tract.ID))
tracts <- gatracts %>%
 left_join(gale, by =c('GEOID' = 'Tract.ID'))
```

### Creating county dataset from full tracts dataset by aggreggating tracts life expectancy and taking the mean to get a mean life expectancy for the county
```{r, message=F}
county <- tracts %>%
  group_by(CTY_NAME) %>%
  summarize(mean_e0 = mean(e.0., na.rm=T))
```

## Life Expectancy Maps
```{r, message=F}

map1 <- tm_shape(tracts) + 
  tm_fill('e.0.',
          style = 'quantile',
         palette= 'BuGn',
         title= 'Life Expectancy, Years')+
    tm_borders(alpha=0.2)+
    tm_layout(main.title='Life Expectancy, GA Tracts',legend.outside=T)+
    tm_credits('Source: U.S. Small-area Life Expectancy Estimates Project, 2010-2015',
               position = c('LEFT', 'BOTTOM'))
    
map2 <- tm_shape(county) + 
  tm_fill('mean_e0',
          style = 'quantile',
         palette= 'Blues',
         title='Mean life Expectancy, Years')+
    tm_borders(alpha=0.2)+
    tm_layout(main.title='Life Expectancy, GA County',legend.outside=T)+
     tm_credits('Source: U.S. Small-area Life Expectancy Estimates Project, 2010-2015',
               position = c('LEFT', 'BOTTOM'))
tmap_arrange(map1,map2, nrow=1, ncol=2)
#Palette was chosen so that the lower value has some color as that is what we might want audience to focus on (not the darker colors). Thought about a diverging palette like RdBu but didn't choose since there isn't a null value for life expectancy and because it made the map harder to read at tract/county level. This might make more sense for the state/country level.
#Chose two different (but similar) palettes since the quantiles are different in both maps
```
## Life Expectancy Ranges by Tract & County

Area | Minimum| Maximum
:----------:|:----------:|:-----------:
County    | 71.8  | 81.0
Tracts | 63.3 | 87.6



```{r,message=F}

tracts_max <- max(tracts$e.0.,na.rm=T) #87.6
tracts_min <- min(tracts$e.0.,na.rm=T) #63.3

county_max <- max(county$mean_e0,na.rm=T) #80.975
county_min <- min(county$mean_e0,na.rm=T) #71.8
#Missing values were removed before finding max/min

```



## 5 Highest Life Expectancy Counties & 5 Lowest Life Expectancy Counties
```{r}
county_test <- county %>%
  arrange(desc(mean_e0))%>%
  slice(1:5,155:159)%>%
  st_drop_geometry()
county_test
```

